Tests of measurement invariance via stochastic processes: {S}ome properties

### 2013-07-30 frontiers/frontiers.Rnw
read in gratitude data and conduct LRT. Update the continuous one. Update the beginning part of the ordinal one.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@226 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

To be more specific, in the following section, we will introduce how to implement multi group analysis in lavaan package. For more details, see {addref}
\subsection{Multiple Group Analysis}
First of all, we need to read in data to R.  
<<read data>>=
# Reading in the data
## (need data here)Data <- read.table("example.dat")
## names(Data) <- c("y1","y2","y3","y4","y5","y6","g")
@
Second, the model need to be specified using lavaan package. 
<<specify model>>=
library(lavaan)
model <- 'f1=~NA*y1+y2+y3
          f2=~NA*y4+y5+y6
          f1 ~~1*f1
          f2 ~~1*f2'
@
Third, we want to test which parameter(s) violate measurement invariance. In lavaan package, loading, manifest variables' intercept and factor means could be tested together or specifically. For more details, please see {addref}. Since in most cases researchers are interested in the factor loading invariance, we will use that as the example in this tutorial.
To implement LRT as ~\ref{equ:LRT}, we need to get estimation for $\hat{\theta}$, without differing groups.

<<theta estimate>>=
# estimate theta 
## theta<-cfa(model,data=Data,group="g",group.equal="loadings")
@
Then we can get estimate for group A and group B. 
<<theta A and theta B estimate>>=
# estimate theta A,B 
## thetaAB<-cfa(model,data=Data,group="g")
@
In the end, we can get the LRT results by calling the measurement invariance function in lavaan. 
<<LRT>>=
# LRT
## measurementInvariance(model,data=Data,group="g",strict=FALSE)
@
\subsection{Extensions for Unknown Group}
The multiple group test performs typically well when the $V$ is a relatively-small number of categories and the sample size is not too large (LRT is sensitive to large sample and will return "false" significance results, details see next section). However, when $V$ (in this case, age) continuous variable, however, multiple-group analysis usually cannot be performed because there are no existing groups. Instead, we can follow the procedure proposed by Ed Merkle and Zeileis. Specifically, we can fit a model whose parameters are restricted to be equal across across all individuals and then examine how individuals' scores $s(\hat{theta};x_{i})$ fluctuate with their values of $V$. To formalize the idea, we assume that the observations are ordered w.r.t. $V$ and define the $k$ dimensional cumulative score process as 
\begin{equation}
\label{equ:B}
B(t;\hat{\theta})=\hat{I}^{-1/2}n^{-1/2}\overset{\lfloor nt \rfloor}{\underset{i=1}{\Sigma}}s(\hat{\theta};x_{i}) (0 \leq t \leq 1)
\end{equation}
where $\lfloor nt \rfloor$ is the integer part of nt and $\hat{I}$ is some consistent estimate of the covariance matrix of the scores,e.g. information matrix. ~\ref{equ:B} simultaneously accounts for the ordering of individuals w.r.t. $V$ and decorrelates the scores between $k$ components of the parameters in the model. 
Further, there exists a functional central limit theorem that allows us to make formal inference with this cumulative score process. Assuming that individuals are independent and the usual ML regularity condition hold, it is possible to show that(addref) 
\begin{equation}
B(\cdot;\hat{\theta})\overset{d}{\rightarrow} B^{0}(\cdot)
\end{equation} 
where $\overset{d}{\rightarrow}$ denotes convergence in distribution and $B^{0}(\cdot)$ is a $k$ dimensional Brownian bridge. Thus, we can construct tests of measurement invariance by comparing the behavior of cumulative score process to that of a Brownian bridge. In practice, there are three test statistics can be used: double maximum, CvM and maxLM. 
\begin{equation}
DM={\underset{i=1,...,n}{\max}}{\underset{j=1,...,k}{\max}}|B(\hat{\theta})_{ij}|
\end{equation}
\begin{equation}
CvM=n^{-1}{\underset{i=1,...,n}{\Sigma}}{\underset{j=1,...,k}{\Sigma}}B(\hat{\theta_{ij}})^{2}
\end{equation}
\begin{equation}
maxLM={\underset{i=1,...,n}{\max}}{\frac{i}{n}(1-\frac{i}{n})}^{-1}{\underset{j=1,...,k}{\Sigma}}B(\hat{\theta}_{ij})^{2})
\end{equation}

Specifically, we will use the previous example, except changing categorical agegroup to continuous age.
Like before, after read in data, we fit the model in lavaan package. 
<<theta estimate>>=
# estimate theta 
## theta<-cfa(model,data=Data,group="g",group.equal="loadings")
@
Then we will use the gefp function in strucchange package to get the $B(\hat{\theta}_{ij})$. Then we can get the three test statistics' significance by using sctest function. 
<<Get B Estimate>>=
library(strucchange)
 for(i in 1:1327) {
  parnum <- c(1,2,3,4,5)
  tname <- c("dmax","CvM","maxLM")
 test <- paste(rep(tname,length(parnum)), rep(parnum, each=length(tname)),
               sep="")
  nrep=1
  pval <- matrix(rep(NA,length(test) * nrep), ncol = length(test))
   colnames(pval) <- test
    mz_gefp <- vector("list", length(parnum))
    for(j in 1:length(parnum)){
      mz_gefp[[j]] <- try(gefp(theta, fit = NULL, vcov = info.mzfit,
                               order.by = yg$agegroup, sandwich = FALSE,
                                 parm = eval(parse(text=parnum[j]))),
                            silent = TRUE)
     if(!inherits(mz_gefp[[j]], "try-error")) {
       pval[i, (j-1)*3 + 1] <- sctest(mz_gefp[[j]],  functional = maxBB)$p.value
      pval[i, (j-1)*3 + 2] <- sctest(mz_gefp[[j]],  functional = meanL2BB)$p.value
       pval[i, (j-1)*3 + 3] <- sctest(mz_gefp[[j]],  functional = supLM(0.1))$p.value
         }
    }
   }
  size = 0.05
  rval <- colMeans(pval < size, na.rm = TRUE)
  return(pval)
@

>>>>

To be more specific, in the following section, we will introduce how to implement multi group analysis in lavaan package. For more details, see {addref}
\subsection{Multiple Group Analysis}
Froh et al had a large sample of youth(n=1401, ranging from late childhood(10 years old) to late adolescent(19 years old) complete the three most widely used adult gratitude inventories, including Gratitude Questionnaire 6(GQ-6; addref), Gratitude Adjective Checklist(GAC; addref) and Gratitude, Resentment, Appreciation Test-Short Form (GRAT-Short Form); addref). In this specific analysis, the authors were interested in whether the youth factor structure for GQ-6 is tau equivalent model or a congeneric model.

To conduct the analysis, first of all, we need to read in data to R and clean the data.
<<read data>>=
# Reading in the data
data("YouthGratitude",package="psychotools")
# remove cases with 'imputed' values(not in 1,...,9)
yg<-YouthGratitude[apply(YouthGratitude[,4:28],1,function(x)all(x %in% 1:9)),]
@
Second, we want to test which parameter(s) violate measurement invariance. In lavaan package, loading, manifest variables' intercept and factor means could be tested together or specifically. For more details, please see {addref}. Since in most cases researchers are interested in the factor loading invariance, we will use that as the example in this tutorial.
To implement LRT as ~\ref{equ:LRT}, we need to get estimation for $\hat{\theta}$, without differing groups.

<<theta estimate>>=
# estimate theta 
library(lavaan)
theta<-cfa(
  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
  data=yg,group="agegroup",meanstructure=TRUE,
  group.equal="loadings")
@
Then we can get estimate for group A and group B. 
<<theta A and theta B estimate>>=
# estimate theta A,B 
thetaAB<-cfa(
  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
  data=yg,group="agegroup",meanstructure=TRUE)
@
In the end, we can get the LRT results by calling the measurement invariance function in lavaan. 
<<LRT>>=
# LRT
LRT<-anova(theta, thetaAB)
LRT
@
\subsection{Extensions for Unknown Group}
The multiple group test performs typically well when the $V$ is a relatively-small number of categories and the sample size is not too large (LRT is sensitive to large sample and will return "false" significance results, details see next section). However, when $V$ (in this case, age) continuous variable, however, multiple-group analysis usually cannot be performed because there are no existing groups. Instead, we can follow the procedure proposed by Ed Merkle and Zeileis. Specifically, we can fit a model whose parameters are restricted to be equal across across all individuals and then examine how individuals' scores $s(\hat{theta};x_{i})$ fluctuate with their values of $V$. To formalize the idea, we assume that the observations are ordered w.r.t. $V$ and define the $k$ dimensional cumulative score process as 
\begin{equation}
\label{equ:B}
B(t;\hat{\theta})=\hat{I}^{-1/2}n^{-1/2}\overset{\lfloor nt \rfloor}{\underset{i=1}{\Sigma}}s(\hat{\theta};x_{i}) (0 \leq t \leq 1)
\end{equation}
where $\lfloor nt \rfloor$ is the integer part of nt and $\hat{I}$ is some consistent estimate of the covariance matrix of the scores,e.g. information matrix. ~\ref{equ:B} simultaneously accounts for the ordering of individuals w.r.t. $V$ and decorrelates the scores between $k$ components of the parameters in the model. 
Further, there exists a functional central limit theorem that allows us to make formal inference with this cumulative score process. Assuming that individuals are independent and the usual ML regularity condition hold, it is possible to show that(addref) 
\begin{equation}
B(\cdot;\hat{\theta})\overset{d}{\rightarrow} B^{0}(\cdot)
\end{equation} 
where $\overset{d}{\rightarrow}$ denotes convergence in distribution and $B^{0}(\cdot)$ is a $k$ dimensional Brownian bridge. Thus, we can construct tests of measurement invariance by comparing the behavior of cumulative score process to that of a Brownian bridge. In practice, there are three test statistics can be used: double maximum, CvM and maxLM. 
\begin{equation}
DM={\underset{i=1,...,n}{\max}}{\underset{j=1,...,k}{\max}}|B(\hat{\theta})_{ij}|
\end{equation}
\begin{equation}
CvM=n^{-1}{\underset{i=1,...,n}{\Sigma}}{\underset{j=1,...,k}{\Sigma}}B(\hat{\theta_{ij}})^{2}
\end{equation}
\begin{equation}
maxLM={\underset{i=1,...,n}{\max}}{\frac{i}{n}(1-\frac{i}{n})}^{-1}{\underset{j=1,...,k}{\Sigma}}B(\hat{\theta}_{ij})^{2})
\end{equation}

Specifically, we will use the previous example, except changing categorical agegroup to continuous age.
Like before, after read in data, we fit the model in lavaan package. 
<<theta estimate>>=
library(lavaan)
theta<-cfa(
  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
  data=yg,group="agegroup",meanstructure=TRUE,
  group.equal="loadings")
@
Then we will use the gefp function in strucchange package to get the $B(\hat{\theta}_{ij})$. Then we can get the three test statistics' significance by using sctest function. 
<<Get >>=
gefp1<-gefp(theta,fit=NULL,order.by=as.numeric(yg$agegroup),
       vcov=info_full,sandwich=FALSE,parm=1:4)
pval=rep(NA,3)
pval[1]<-sctest(gefp1,functional=maxBB)$p.value
pval[2]<-sctest(gefp1,functional=meanL2BB)$p.value
pval[3]<-sctest(gefp1,functional=supLM(0.1))$p.value
pval
@

\subsection{Ordinal Test}
However, in this dataset, $V$ is actually ordinal instead of continuous. There is only a partial ordering of individuals with respect to $V$,i.e. observations with the same level of $V$ have no unique ordering. The ordinal statistics proposed here are similar to those described in Equation and above.
